---
title: "The Base Plotting System in ggplot2"
output: html_notebook
---

## Example Dataset

```{r}
library(ggplot2)
str(mpg)
```




## ggplot2 "Hello, world!"

```{r}
qplot(displ, hwy, data = mpg)
```


## Modifying aesthetics

```{r}
qplot(displ, hwy, data = mpg, color = drv)

```



## Adding a geom

```{r}
qplot(displ, hwy, data = mpg, geom = c("point", "smooth"))

```


## Histograms

```{r}
qplot(hwy, data = mpg, fill = drv)

```



## Facets

```{r, fig.width=4.5}
qplot(displ, hwy, data = mpg, facets = . ~ drv)
qplot(hwy, data = mpg, facets = drv ~ ., binwidth = 2)
```

---

## MAACS Cohort

- Mouse Allergen and Asthma Cohort Study
- Baltimore children (aged 5-17)
- Persistent asthma, exacerbation in past year
- Study indoor environment and its relationship with asthma morbidity
- Recent publication: http://goo.gl/WqE9j8

```{r,echo=FALSE}
eno <- read.csv("eno.csv")
skin <- read.csv("skin.csv")
env <- read.csv("environmental.csv")
m <- merge(eno, env, by = "id")
maacs <- merge(m, skin, by = "id")
```

## Example: MAACS

```{r}
str(maacs)
```


## Histogram of eNO

```{r}
qplot(log(eno), data = maacs)
```


## Histogram by Group

```{r}
qplot(log(eno), data = maacs, fill = mopos)
```


## Density Smooth

```{r, fig.width=4.5}
qplot(log(eno), data = maacs, geom = "density")
qplot(log(eno), data = maacs, geom = "density", color = mopos)
```


## Scatterplots: eNO vs. PM$_{2.5}$

```{r, fig.width=2.7}
qplot(log(pm25), log(eno), data = maacs)
qplot(log(pm25), log(eno), data = maacs, shape = mopos)
qplot(log(pm25), log(eno), data = maacs, color = mopos)
```



## Scatterplots: eNO vs. PM$_{2.5}$

```{r}
qplot(log(pm25), log(eno), data = maacs, color = mopos, 
      geom = c("point", "smooth"), method = "lm")
```



## Scatterplots: eNO vs. PM$_{2.5}$

```{r, fig.width=9}
qplot(log(pm25), log(eno), data = maacs, geom = c("point", "smooth"), 
      method = "lm", facets = . ~ mopos)
```


## Summary of qplot()

- The `qplot()` function is the analog to `plot()` but with many built-in features
- Syntax somewhere in between base/lattice
- Produces very nice graphics, essentially publication ready (if you like the design)
- Difficult to go against the grain/customize (don’t bother; use full ggplot2 power in that case)


## Resources

- The _ggplot2_ book by Hadley Wickham
- The _R Graphics Cookbook_ by Winston Chang (examples in base plots and in ggplot2)
- ggplot2 web site (http://ggplot2.org)
- ggplot2 mailing list (http://goo.gl/OdW3uB), primarily for developers
