---
title: "Noncompartmental Analysis"
output: html_notebook
---

### Read in data
```{r}
data1<-read.csv("finalPKdata.csv", header=TRUE)
```


### Activate required libraries
```{r}
list.of.packages <- c("lattice", 
                      "PKNCA", 
                      "ggplot2", 
                      "knitr", 
                      "nlme", 
                      "reshape2")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
lapply(list.of.packages, require, character.only = TRUE)
# this should return 5 TRUE statements (one for each package)
# you can also lapply library instead of require
```


### Plot trial demographic data

#### Histogram of age
```{r}
hist(data1$Age)
hist(data1$Age,cex=2.0, cex.lab=1.5, cex.axis=1.5,xlab="Age (yrs)", 
     xlim=c(30,65),main="Age of Patients")
```


#### Histogram of body weight
```{r}
hist(data1$Weight,cex=2.0, cex.lab=1.5, cex.axis=1.5,xlab="Weight (kgs)", 
     xlim=c(50,100),main="Body Weight of Patients")
```


#### Histogram of Serum Creatinine
```{r}
hist(data1$SCreatinine,cex=2.0, cex.lab=1.5, cex.axis=1.5,xlab="SerCreatinine (units))", 
     xlim=c(0,2),main="Serum Creatinine of Patients")
```


#### Histogram of Serum ALT
```{r}
hist(data1$SerumALT,cex=2.0, cex.lab=1.5, cex.axis=1.5,xlab="Serum ALT (units))", 
     xlim=c(10,25),main="Serum ALT of Patients")
```


### PK Analysis

#### Basic plot

```{r}
plot(data1$Conc~data1$Time, ylab="Concentration (ng/mL)", 
     xlab="Time (hr)")
```


#### group data and plot
```{r}
data1.2<-groupedData(Conc~Time|ID, data=data1, labels = list(x="Time(hr)", 
                y="Conc(ng/mL)"))
plot(data1.2, aspect=1/1, ylab="Concentration (ng/mL)", xlab="Time (hr)")
data1.2$logConc<-log(data1.2$Conc)

data1.2log<-groupedData(logConc~Time|ID, data=data1.2, labels = list(x="Time(hr)", 
      y="Log Conc(ng/mL)"))
plot(data1.2log, aspect=1/1, ylab="Log Concentration (ng/mL)", xlab="Time (hr)")

```

#### PK Profiles by Sex
```{r}
data1.3<-groupedData(Conc~Time|Gender, data=data1, labels=list(x="Time", 
            y="Concentration"))
plot(data1.3, aspect=1/1, ylab="Concentration (ng/mL)", xlab="Time (hr)")
```


#### Log transform dependent variable
```{r}
data1$logConc<-log(data1$Conc)
data1.3log<-groupedData(logConc~Time|Gender, data=data1, labels=list(x="Time", 
        y="Log Concentration"))
plot(data1.3log, aspect=1/1, ylab="Log Concentration (ng/mL)", 
     xlab="Time (hr)")
```


### Perform NCA

#### By default it is groupedData; convert it to a data frame for use
#### maps conc to any variable or grouping factor
```{r}
my.conc <- PKNCAconc(as.data.frame(data1), Conc~Time|ID)
```


#### Dosing data needs to only have one row per dose, so subset for that first
```{r}
d.dose <- unique(data1[data1$Time == 0,
   c("ID","Time","Dose", "Conc", "Age","Weight","SCreatinine","SerumALT",
     "Gender","Race")])

knitr::kable(d.dose,
             caption="Dosing data extracted from data set")
```


#### By dose
```{r}
my.dose <- PKNCAdose(d.dose, Dose~Time|ID)
my.dose
```


#### Combine dose and conc data
```{r}
my.data.automatic <- PKNCAdata(my.conc, my.dose)
summary(my.data.automatic)
knitr::kable(PKNCA.options("single.dose.aucs"))
knitr::kable(my.data.automatic$intervals)
```


#### Compute parameters
```{r}
my.results.automatic <- pk.nca(my.data.automatic)
summary(my.results.automatic)
```


#### Specify start/stop times
```{r}
my.intervals <- data.frame(start=0,
                           end=Inf,
                           cmax=TRUE,
                           tmax=TRUE,
                           aucinf.obs=TRUE,
                           auclast=FALSE,
                           cl.obs=TRUE,
                           vz.obs=TRUE,
                           half.life=TRUE)
my.data.manual <- PKNCAdata(my.conc, my.dose,intervals=my.intervals)

knitr::kable(my.data.manual$interval)
summary(my.data.manual)

my.results.manual <- pk.nca(my.data.manual)
summary(my.results.manual)
my.results.manual$result

write.csv(my.results.manual$result, file="PKNCAdata.csv")

knitr::kable(summary(my.results.manual))

```

#### Make a data.frame for PK manual results
```{r}
PKresults<-my.results.manual$result
```

#### Drop first two columns in PK results
```{r}
PKresults1<-PKresults[c(-1, -2)]
```


#### Subset out only t=0hr rows in main data.frame
```{r}
data0<-data1[data1$Time==0,]
```

#### Drop the log conc column

```{r}
data0.1<-data0[c(-11)]
PKresults2<-t(PKresults1)
```

#### Look up melt documentation
```{r}
?melt
```


#### Melt PK data
```{r}
mPKresults<-melt(PKresults1, id=c("ID", "PPTESTCD"))
```

#### Look up cast documentation
```{r}
?cast
```

#### Cast the melted data ###
```{r}
        ### cast(data, formula, function) ###
PKresults3<-dcast(mPKresults, ID~variable+PPTESTCD)
write.csv(PKresults3, file="PKNCAresults.csv")
```


#### Cbind pk results to subsetted data.frame
```{r}
alldata<-cbind2(PKresults3, data0.1)
head(alldata)
write.csv(alldata, file="FinalPKNCA1.csv")
```


### Subset dataset for PK grouping
#### PK by Sex
```{r}
my.conc.sex <- PKNCAconc(as.data.frame(data1), 
        Conc~Time|Gender+ID)
my.dose.sex <- PKNCAdose(d.dose, Dose~Time|Gender+ID)
my.data.auto.sex <- PKNCAdata(my.conc.sex, my.dose.sex)
my.results.auto.sex <- pk.nca(my.data.auto.sex)
summary(my.results.auto.sex)

my.data.manual.sex <- PKNCAdata(my.conc.sex, my.dose.sex,
                            intervals=my.intervals)
my.results.manual.sex <- pk.nca(my.data.manual.sex)
summary(my.results.manual.sex)

```

####  PK by Race
```{r}
my.conc.race <- PKNCAconc(as.data.frame(data1), Conc~Time|Race+ID)
my.dose.race <- PKNCAdose(d.dose, Dose~Time|Race+ID)
my.data.auto.race <- PKNCAdata(my.conc.race, my.dose.race)
my.results.auto.race <- pk.nca(my.data.auto.race)
summary(my.results.auto.race)

my.data.manual.race <- PKNCAdata(my.conc.race, my.dose.race,
                                intervals=my.intervals)
my.results.manual.race <- pk.nca(my.data.manual.race)
summary(my.results.manual.race)
```

###  PK Correlations with Patient Demographics

#### Look at pairs documentation
```{r}
?pairs
```

#### Scatter plot matrix for broad overview
```{r}
# Convert PPORRES class to numeric
class(alldata$PPORRES_cmax)<-"numeric"
class(alldata$PPORRES_aucinf.obs)<-"numeric"

# Basic Scatterplot Matrix for exposure vs continuous variables only
pairs(~PPORRES_cmax+PPORRES_aucinf.obs +Age + Weight +
        SCreatinine+SerumALT,data=alldata, 
      main="Scatterplot Matrix for exposure vs continuous variables only")
write.csv(alldata, file="alldata1.csv")
alldata1<-read.csv("alldata1.csv", header=TRUE)

pairs(~PPORRES_cmax+PPORRES_aucinf.obs+Race+Gender,data=alldata1, 
      main="Scatterplot Matrix for exposure vs ordinal variables only")

pairs(~PPORRES_cl.obs+Age+Weight+SCreatinine+SerumALT,data=alldata1, 
      main="Scatterplot Matrix for clearance vs continuous variables only")

```

#### CL by WT
```{r}
plot(alldata$Weight, alldata$PPORRES_cl.obs)
plot.clwt<- ggplot() + geom_point(data = alldata1, 
  aes(x=Weight, y = PPORRES_cl.obs), size=2.5, colour='blue') +
  xlab('Weight (kg)') +
  ylab('Clearance(L/hr)')
plot.clwt+ theme(panel.background = element_rect(fill='grey', colour='red'))+
  theme(axis.title.y = element_text(colour = 'black', size = 20))+
  theme(axis.text.y = element_text(size = 15))+
  theme(axis.title.x = element_text(colour = 'black', size = 20))+
  theme(axis.text.x = element_text(size = 15))+
  ggtitle("Drug Clearance by Body Weight")+
  theme(plot.title = element_text(lineheight=.8, face="bold",size = 18))+
  theme(legend.position="top")+
  theme(legend.title=element_blank())+
  theme(legend.text = element_text(colour="blue", size = 12, face = "bold"))
```


#### CL by Age
```{r}
plot(alldata$Age, alldata$PPORRES_cl.obs)
plot.clage<- ggplot() + geom_point(data = alldata, aes(x=Age, y = PPORRES_cl.obs), size=2.5, colour='blue') +
  xlab('Age(yr)') +
  ylab('Clearance(L/hr)')
plot.clage+ theme(panel.background = element_rect(fill='grey', colour='red'))+
  theme(axis.title.y = element_text(colour = 'black', size = 20))+
  theme(axis.text.y = element_text(size = 15))+
  theme(axis.title.x = element_text(colour = 'black', size = 20))+
  theme(axis.text.x = element_text(size = 15))+
  ggtitle("Drug Clearance by Age")+
  theme(plot.title = element_text(lineheight=.8, face="bold",size = 18))+
  theme(legend.position="top")+
  theme(legend.title=element_blank())+
  theme(legend.text = element_text(colour="blue", size = 12, face = "bold"))


par(mfrow=c(1,2))
plot(alldata$Weight, alldata$PPORRES_cl.obs)
plot(alldata$Age, alldata$PPORRES_cl.obs)

```

#### CL by Gender
```{r}
plot(alldata1$Gender, alldata1$PPORRES_cl.obs)
plot.clsex<- ggplot() + geom_boxplot(data = alldata, aes(x=Gender, y = PPORRES_cl.obs), size=1.5, colour='blue') +
  xlab('Sex)') +
  ylab('Clearance(L/hr)')
plot.clsex+ theme(panel.background = element_rect(fill='grey', colour='red'))+
  theme(axis.title.y = element_text(colour = 'black', size = 20))+
  theme(axis.text.y = element_text(size = 15))+
  theme(axis.title.x = element_text(colour = 'black', size = 20))+
  theme(axis.text.x = element_text(size = 15))+
  ggtitle("Drug Clearance by Sex")+
  theme(plot.title = element_text(lineheight=.8, face="bold",size = 18))+
  theme(legend.position="top")+
  theme(legend.title=element_blank())+
  theme(legend.text = element_text(colour="blue", size = 12, face = "bold"))

```


#### CL by Race
```{r}
par(mfrow=c(1,1))
plot(alldata1$Race, alldata1$PPORRES_cl.obs)
plot.clrace<- ggplot() + geom_boxplot(data = alldata, aes(x=Race, y = PPORRES_cl.obs), size=1.5, colour='blue') +
  xlab('Race') +
  ylab('Clearance(L/hr)')
plot.clrace+ theme(panel.background = element_rect(fill='grey', colour='red'))+
  theme(axis.title.y = element_text(colour = 'black', size = 20))+
  theme(axis.text.y = element_text(size = 15))+
  theme(axis.title.x = element_text(colour = 'black', size = 20))+
  theme(axis.text.x = element_text(size = 15))+
  ggtitle("Drug Clearance by Race")+
  theme(plot.title = element_text(lineheight=.8, face="bold",size = 18))+
  theme(legend.position="top")+
  theme(legend.title=element_blank())+
  theme(legend.text = element_text(colour="blue", size = 12, face = "bold"))

par(mfrow=c(1,2))
plot(alldata$Gender, alldata$PPORRES_cl.obs)
plot(alldata$Race, alldata$PPORRES_cl.obs)


par(mfrow=c(1,1))

```

#### Cmax by WT
```{r}
plot(alldata$Weight, alldata$PPORRES_cmax)
plot.cmaxwt<- ggplot() + geom_point(data = alldata, aes(x=Weight, y = PPORRES_cmax), size=2.5, colour='blue') +
  xlab('Weight (kg)') +
  ylab('Cmax (ng/mL))')
plot.cmaxwt+ theme(panel.background = element_rect(fill='grey', colour='red'))+
  theme(axis.title.y = element_text(colour = 'black', size = 20))+
  theme(axis.text.y = element_text(size = 15))+
  theme(axis.title.x = element_text(colour = 'black', size = 20))+
  theme(axis.text.x = element_text(size = 15))+
  ggtitle("Cmax by Body Weight")+
  theme(plot.title = element_text(lineheight=.8, face="bold",size = 18))+
  theme(legend.position="top")+
  theme(legend.title=element_blank())+
  theme(legend.text = element_text(colour="blue", size = 12, face = "bold"))

```


#### Cmax by Age
```{r}
plot(alldata$Age, alldata$PPORRES_cmax)
plot.cmaxage<- ggplot() + geom_point(data = alldata, aes(x=Age, y = PPORRES_cmax), size=2.5, colour='blue') +
  xlab('Age(yr)') +
  ylab('Cmax (ng/mL)')
plot.cmaxage+ theme(panel.background = element_rect(fill='grey', colour='red'))+
  theme(axis.title.y = element_text(colour = 'black', size = 20))+
  theme(axis.text.y = element_text(size = 15))+
  theme(axis.title.x = element_text(colour = 'black', size = 20))+
  theme(axis.text.x = element_text(size = 15))+
  ggtitle("Cmax by Age")+
  theme(plot.title = element_text(lineheight=.8, face="bold",size = 18))+
  theme(legend.position="top")+
  theme(legend.title=element_blank())+
  theme(legend.text = element_text(colour="blue", size = 12, face = "bold"))


par(mfrow=c(1,2))
plot(alldata$Weight, alldata$PPORRES_cmax)
plot(alldata$Age, alldata$PPORRES_cmax)

```



#### Cmax by Gender
```{r}
plot(alldata$Gender, alldata$PPORRES_cmax)
plot.clsex<- ggplot() + geom_boxplot(data = alldata, aes(x=Gender, y = PPORRES_cmax), size=1.5, colour='blue') +
  xlab('Sex)') +
  ylab('Cmax (ng/mL)')
plot.clsex+ theme(panel.background = element_rect(fill='grey', colour='red'))+
  theme(axis.title.y = element_text(colour = 'black', size = 20))+
  theme(axis.text.y = element_text(size = 15))+
  theme(axis.title.x = element_text(colour = 'black', size = 20))+
  theme(axis.text.x = element_text(size = 15))+
  ggtitle("Cmax by Sex")+
  theme(plot.title = element_text(lineheight=.8, face="bold",size = 18))+
  theme(legend.position="top")+
  theme(legend.title=element_blank())+
  theme(legend.text = element_text(colour="blue", size = 12, face = "bold"))

```


#### Cmax by Race
```{r}
par(mfrow=c(1,1))
plot(alldata$Race, alldata$PPORRES_cmax)
plot.clrace<- ggplot() + geom_boxplot(data = alldata, aes(x=Race, y = PPORRES_cmax), size=1.5, colour='blue') +
  xlab('Race') +
  ylab('Cmax (ng/mL))')
plot.clrace+ theme(panel.background = element_rect(fill='grey', colour='red'))+
  theme(axis.title.y = element_text(colour = 'black', size = 20))+
  theme(axis.text.y = element_text(size = 15))+
  theme(axis.title.x = element_text(colour = 'black', size = 20))+
  theme(axis.text.x = element_text(size = 15))+
  ggtitle("Cmax by Race")+
  theme(plot.title = element_text(lineheight=.8, face="bold",size = 18))+
  theme(legend.position="top")+
  theme(legend.title=element_blank())+
  theme(legend.text = element_text(colour="blue", size = 12, face = "bold"))

par(mfrow=c(1,2))
plot(alldata$Gender, alldata$PPORRES_cmax)
plot(alldata$Race, alldata$PPORRES_cmax)


par(mfrow=c(1,1))

```

#### AUCinf by WT
```{r}
plot(alldata$Weight, alldata$PPORRES_aucinf.obs)
plot.aucinfwt<- ggplot() + geom_point(data = alldata, aes(x=Weight, y = PPORRES_aucinf.obs), size=2.5, colour='blue') +
  xlab('Weight (kg)') +
  ylab('AUCinf (hr*ng/mL))')
plot.aucinfwt+ theme(panel.background = element_rect(fill='grey', colour='red'))+
  theme(axis.title.y = element_text(colour = 'black', size = 20))+
  theme(axis.text.y = element_text(size = 15))+
  theme(axis.title.x = element_text(colour = 'black', size = 20))+
  theme(axis.text.x = element_text(size = 15))+
  ggtitle("AUCinf by Body Weight")+
  theme(plot.title = element_text(lineheight=.8, face="bold",size = 18))+
  theme(legend.position="top")+
  theme(legend.title=element_blank())+
  theme(legend.text = element_text(colour="blue", size = 12, face = "bold"))

```

#### AUCinf by Age
```{r}
plot(alldata$Age, alldata$PPORRES_aucinf.obs)
plot.aucinfage<- ggplot() + geom_point(data = alldata, aes(x=Age, y = PPORRES_aucinf.obs), size=2.5, colour='blue') +
  xlab('Age(yr)') +
  ylab('AUCinf (hr*ng/mL)')
plot.aucinfage+ theme(panel.background = element_rect(fill='grey', colour='red'))+
  theme(axis.title.y = element_text(colour = 'black', size = 20))+
  theme(axis.text.y = element_text(size = 15))+
  theme(axis.title.x = element_text(colour = 'black', size = 20))+
  theme(axis.text.x = element_text(size = 15))+
  ggtitle("AUCinf by Age")+
  theme(plot.title = element_text(lineheight=.8, face="bold",size = 18))+
  theme(legend.position="top")+
  theme(legend.title=element_blank())+
  theme(legend.text = element_text(colour="blue", size = 12, face = "bold"))


par(mfrow=c(1,2))
plot(alldata$Weight, alldata$PPORRES_aucinf.obs)
plot(alldata$Age, alldata$PPORRES_aucinf.obs)
```

##### AUCinf by Gender
```{r}
plot(alldata$Gender, alldata$PPORRES_aucinf.obs)
plot.aucinfsex<- ggplot() + geom_boxplot(data = alldata, aes(x=Gender, y = PPORRES_aucinf.obs), size=1.5, colour='blue') +
  xlab('Sex)') +
  ylab('AUCinf (hr*ng/mL)')
plot.aucinfsex+ theme(panel.background = element_rect(fill='grey', colour='red'))+
  theme(axis.title.y = element_text(colour = 'black', size = 20))+
  theme(axis.text.y = element_text(size = 15))+
  theme(axis.title.x = element_text(colour = 'black', size = 20))+
  theme(axis.text.x = element_text(size = 15))+
  ggtitle("AUCinf by Sex")+
  theme(plot.title = element_text(lineheight=.8, face="bold",size = 18))+
  theme(legend.position="top")+
  theme(legend.title=element_blank())+
  theme(legend.text = element_text(colour="blue", size = 12, face = "bold"))
```

##### AUCinf by Race
```{r}
par(mfrow=c(1,1))
plot(alldata$Race, alldata$PPORRES_aucinf.obs)
plot.aucinfrace<- ggplot() + geom_boxplot(data = alldata, aes(x=Race, y = PPORRES_aucinf.obs), size=1.5, colour='blue') +
  xlab('Race') +
  ylab('AUCinf (hr*ng/mL)')
plot.aucinfrace+ theme(panel.background = element_rect(fill='grey', colour='red'))+
  theme(axis.title.y = element_text(colour = 'black', size = 20))+
  theme(axis.text.y = element_text(size = 15))+
  theme(axis.title.x = element_text(colour = 'black', size = 20))+
  theme(axis.text.x = element_text(size = 15))+
  ggtitle("AUCinf by Race")+
  theme(plot.title = element_text(lineheight=.8, face="bold",size = 18))+
  theme(legend.position="top")+
  theme(legend.title=element_blank())+
  theme(legend.text = element_text(colour="blue", size = 12, face = "bold"))

par(mfrow=c(1,2))
plot(alldata$Gender, alldata$PPORRES_aucinf.obs)
plot(alldata$Race, alldata$PPORRES_aucinf.obs)
```
