---
title: "The Base Plotting System in R"
output: html_notebook
---

# About
This material is adapted from Roger Peng's [Exploratory Data Analysis Course](https://www.coursera.org/learn/exploratory-data-analysis).

You can see Roger's original [source code](https://github.com/DataScienceSpecialization/courses/tree/master/04_ExploratoryAnalysis) on GitHub.

I also recommend looking at the [repo for Roger's book](https://github.com/rdpeng/exdata), which is available for little or no money (you decide!) on [Leanpub](https://leanpub.com/exdata/).

## Simple Base Graphics: Histogram

```{r,fig.height=5}
library(datasets)
hist(airquality$Ozone)  ## Draw a new plot
```



## Simple Base Graphics: Scatterplot

```{r,fig.height=5}
library(datasets)
with(airquality, plot(Wind, Ozone))
```



## Simple Base Graphics: Boxplot

```{r,fig.height=5}
library(datasets)
airquality <- transform(airquality, Month = factor(Month))
boxplot(Ozone ~ Month, airquality, xlab = "Month", ylab = "Ozone (ppb)")
```



## Some Important Base Graphics Parameters

Many base plotting functions share a set of parameters. Here are a few
key ones:

- `pch`: the plotting symbol (default is open circle)

- `lty`: the line type (default is solid line), can be dashed, dotted, etc.

- `lwd`: the line width, specified as an integer multiple

- `col`: the plotting color, specified as a number, string, or hex
  code; the `colors()` function gives you a vector of colors by name

- `xlab`: character string for the x-axis label

- `ylab`: character string for the y-axis label




## Some Important Base Graphics Parameters

The `par()` function is used to specify *global* graphics parameters
that affect all plots in an R session. These parameters can be
overridden when specified as arguments to specific plotting functions.

- `las`: the orientation of the axis labels on the plot
- `bg`: the background color
- `mar`: the margin size
- `oma`: the outer margin size (default is 0 for all sides)
- `mfrow`: number of plots per row, column (plots are filled row-wise) 
- `mfcol`: number of plots per row, column (plots are filled column-wise)



## Some Important Base Graphics Parameters

Default values for global graphics parameters

```{r}
par("lty")
par("col")
par("pch")
```



## Some Important Base Graphics Parameters

Default values for global graphics parameters

```{r}
par("bg")
par("mar")
par("mfrow")
```



## Base Plotting Functions

- `plot`: make a scatterplot, or other type of plot depending on the
  class of the object being plotted

- `lines`: add lines to a plot, given a vector x values and a
  corresponding vector of y values (or a 2-column matrix); this
  function just connects the dots

- `points`: add points to a plot
- `text`: add text labels to a plot using specified x, y coordinates
- `title`: add annotations to x, y axis labels, title, subtitle, outer margin 
- `mtext`: add arbitrary text to the margins (inner or outer) of the plot 
- `axis`: adding axis ticks/labels



## Base Plot with Annotation

```{r,fig.height=5}
library(datasets)
with(airquality, plot(Wind, Ozone))
title(main = "Ozone and Wind in New York City")  ## Add a title
```



## Base Plot with Annotation

```{r,fig.height=5}
with(airquality, plot(Wind, Ozone, main = "Ozone and Wind in New York City"))
with(subset(airquality, Month == 5), points(Wind, Ozone, col = "blue"))
```



## Base Plot with Annotation

```{r,fig.height=5}
with(airquality, plot(Wind, Ozone, main = "Ozone and Wind in New York City", type = "n"))
with(subset(airquality, Month == 5), points(Wind, Ozone, col = "blue"))
with(subset(airquality, Month != 5), points(Wind, Ozone, col = "red"))
legend("topright", pch = 1, col = c("blue", "red"), legend = c("May", "Other Months"))
```



## Base Plot with Regression Line

```{r,fig.height=5}
with(airquality, plot(Wind, Ozone, main = "Ozone and Wind in New York City", pch = 20))
model <- lm(Ozone ~ Wind, airquality)
abline(model, lwd = 2)
```



## Multiple Base Plots

```{r,fig.height=5,fig.width=14}
par(mfrow = c(1, 2))
with(airquality, {
	plot(Wind, Ozone, main = "Ozone and Wind")
	plot(Solar.R, Ozone, main = "Ozone and Solar Radiation")
})
```



## Multiple Base Plots

```{r,fig.height=4,fig.width=12}
par(mfrow = c(1, 3), mar = c(4, 4, 2, 1), oma = c(0, 0, 2, 0))
with(airquality, {
	plot(Wind, Ozone, main = "Ozone and Wind")
	plot(Solar.R, Ozone, main = "Ozone and Solar Radiation")
	plot(Temp, Ozone, main = "Ozone and Temperature")
	mtext("Ozone and Weather in New York City", outer = TRUE)
})
```



## Summary

* Plots in the base plotting system are created by calling successive R functions to "build up" a plot

* Plotting occurs in two stages:
  - Creation of a plot
  - Annotation of a plot (adding lines, points, text, legends)

* The base plotting system is very flexible and offers a high degree of control over plotting
