---
title: "Linear Regression"
output: html_notebook
---

#### Read in and explore data
```{r}
data<-read.csv("SimpleLinReg_day2.csv", header=TRUE)
str(data)
head(data)
```
#### Plot dose versus response
```{r}
dev.off
plot(data$dose, data$resp)
```

## LINEAR REGRESSION

#### Perform Linear Regression
```{r}
attach(data)
myfit<-lm(resp~dose)
summary(myfit)

resid(myfit)
fitted(myfit)
plot(dose,resp,cex = 1.5,cex.axis=1.5,cex.lab=1.5,ylab=
       "Change from baseline in DBP (mmHg)",
     xlab="Dose(mg)",pch = 16, 
     main=" Response vs Dose")

lines(dose, fitted(myfit),lwd = 3.0,col="blue")
```
#### Calculate confidence intervals
```{r}
confint(myfit)
```
#### Perform Analysis of Variance (ANOVA)
```{r}
anova(myfit)
```


#### Multiple Comparisons using Tukey's Method
```{r}
TukeyHSD(aov(resp~as.factor(dose)))
detach(data)
```


## MULTIPLE LINEAR REGRESSION
#### Read in, explore, and attach data
```{r}
fudata<-read.csv("MltplLinReg_Day2.csv", header = TRUE)
str(fudata)
head(fudata)
attach(fudata)
```

#### Exploratory plots
```{r}
par(mfrow=c(1,2))
hist(FU_CL, breaks=10, xlab="5-FU Clearance (L/min)", 
     cex=1.5, cex.axis=1.5)
qqnorm(FU_CL, pch=16, cex=1.5, cex.axis=1.5, 
       ylab = "5-FU Clearance Quantiles")
qqline(FU_CL, col="blue", cex=1.5, lwd=2.0)
```


#### Clearance is log-normally distributed
```{r}
fudata$logFUCL<-log(FU_CL)
attach(fudata)
par(mfrow=c(1,2))
hist(logFUCL, breaks=10, xlab="5-FU Clearance (L/min)", 
     cex=1.5, cex.axis=1.5)
qqnorm(logFUCL, pch=16, cex=1.5, cex.axis=1.5, 
       ylab = "5-FU Clearance Quantiles")
qqline(logFUCL, col="blue", cex=1.5, lwd=2.0)

```



#### Semi-log plots
```{r}
par(mfrow=c(2,3))
plot(AGE,logFUCL,ylim=c(-1,1),cex = 1.5,
     cex.axis=2.0,cex.lab=2.0,
     ylab="5-FU clearance, L/min",xlab="Age,yrs",pch = 16)
clfit1<-lm(logFUCL~AGE)
lines(AGE, fitted(clfit1),lwd = 2.0,col="blue")

boxplot(logFUCL~SEX,cex=1.5,cex.axis=2.0,cex.lab=2.0,
        names=c("Male","Female"))
boxplot(logFUCL~MTX,cex=1.5,cex.axis=2.0,cex.lab=2.0,
        names=c("No MTX","MTX"))
plot(BSA,logFUCL,ylim=c(-1,1),cex = 1.5,cex.axis=2.0,
     cex.lab=2.0,
     ylab="5-FU clearance, L/min",xlab="BSA,m2",pch = 16)
clfit2<-lm(logFUCL~BSA)
lines(BSA, fitted(clfit2),lwd = 2.0,col="red")

plot(DOSE,logFUCL,ylim=c(-1,1),cex = 1.5,cex.axis=2.0,cex.lab=2.0,
     ylab="5-FU clearance, L/min",xlab="DOSE,mg",pch = 16)
clfit3<-lm(logFUCL~DOSE)
lines(DOSE, fitted(clfit3),lwd = 2.0,col="green")

```


#### Relationship Among Covariates
```{r}
pairs(fudata[,2:6], pch=16, cex=1.5)
  ## correlation matrix##
cor(fudata[,2:6])
cor.test(MTX,DOSE)
cor.test(AGE,BSA)
cor.test(SEX,BSA)
```

#### Univariate analysis
```{r}
  ### CL vs sex
plot(fudata$SEX, data$FU_CL)
attach(fudata)
LM_fu_sex<-lm(log(FU_CL) ~ factor(SEX))
summary(LM_fu_sex)
resid(LM_fu_sex)
fitted(LM_fu_sex)
plot(SEX,FU_CL,cex = 1.5,cex.axis=1.5,cex.lab=1.5,ylab=
       "5-FU Clearance (L/min))",xlab="SEX",pch = 16)
lines(SEX, fitted(LM_fu_sex),lwd = 3.0,col="blue")

```


#### CL vs Dose
```{r}
plot(fudata$DOSE, data$logFUCL)
attach(fudata)
LM_fu_dose<-lm(logFUCL ~ DOSE)
summary(LM_fu_dose)
resid(LM_fu_dose)
fitted(LM_fu_dose)
plot(DOSE,logFUCL,cex = 1.5,cex.axis=1.5,cex.lab=1.5, ylab=
       "5-FU Clearance (L/min))",xlab="Dose (mg)", pch = 16)
lines(DOSE, fitted(LM_fu_dose),lwd = 3.0,col="blue")

```



#### CL vs Age
```{r}

plot(fudata$AGE, data$FU_CL)
attach(fudata)
LM_fu_age<-lm(log(FU_CL) ~ AGE)
summary(LM_fu_age)
resid(LM_fu_age)
fitted(LM_fu_age)
plot(AGE,logFUCL,cex = 1.5,cex.axis=1.5,cex.lab=1.5, ylab=
       "5-FU Clearance (L/min))",xlab="Age (yrs)", pch = 16)
lines(AGE, fitted(LM_fu_age),lwd = 3.0,col="blue")

```

#### CL vs BSA
```{r}
plot(fudata$BSA, data$FU_CL)
attach(fudata)
LM_fu_bsa<-lm(log(FU_CL) ~ BSA)
summary(LM_fu_bsa)
resid(LM_fu_bsa)
fitted(LM_fu_bsa)
plot(BSA,FU_CL,cex = 1.5,cex.axis=1.5,cex.lab=1.5, ylab=
       "5-FU Clearance (L/min))",xlab="BSA (m**2)", pch = 16)
lines(BSA, fitted(LM_fu_bsa),lwd = 3.0,col="blue")

```


#### CL vs MTX
```{r}
plot(fudata$MTX, data$FU_CL)
attach(fudata)
LM_fu_mtx<-lm(log(FU_CL) ~ factor(MTX))
summary(LM_fu_mtx)
resid(LM_fu_mtx)
fitted(LM_fu_mtx)
plot(MTX,FU_CL,cex = 1.5,cex.axis=1.5,cex.lab=1.5, ylab=
       "5-FU Clearance (L/min))",xlab="MTX", pch = 16)
lines(MTX, fitted(LM_fu_mtx),lwd = 3.0,col="blue")

```



## Scaling of continuous covariates
##### Dose, scaled by 1000
```{r}
fudata$DOSE1<-DOSE/1000
fudata$AGE1<-AGE/50
fudata$BSA1<-BSA/1.7
fudata$logFUCL<-log(FU_CL)
head(fudata)
```
### scaled univariate analysis
#### CL vs Dose

```{r}
attach(fudata)
detach(fudata)
LM_fu_dose2<-lm(fudata$logFUCL ~ fudata$DOSE1)
summary(LM_fu_dose2)

```


#### CL vs Age
```{r}
LM_fu_age2<-lm(log(FU_CL) ~ fudata$AGE1)
summary(LM_fu_age2)
```


#### CL vs BSA
```{r}
LM_fu_bsa2<-lm(log(FU_CL) ~ fudata$BSA1)
summary(LM_fu_bsa2)

```



### Full model
```{r}
full_lm<-lm(fudata$logFUCL ~ fudata$DOSE1 + fudata$AGE1 + 
        fudata$BSA1 + factor(fudata$SEX) + factor(fudata$MTX))
summary(full_lm)
```


#### sig p-value from this F-test says at least one beta is not=0
#### hence reject Null---

#### Backward elimination to help choose estimate
#### remove predictor with large p-value
#### in this case, MTX predictor had highest p-value, so removed

```{r}
red_lm1<-lm(fudata$logFUCL ~ fudata$DOSE1 + fudata$AGE1 + 
        fudata$BSA1 + factor(fudata$SEX))
summary(red_lm1)
```


#### Now dose even more sig, even BSA sig now
#### Now, remove Age b/c had highest p-value

```{r}
red_lm2<-lm(fudata$logFUCL ~ fudata$DOSE1 + fudata$BSA1 + 
          factor(fudata$SEX))
summary(red_lm2)
```


#### Final model
```{r}
red_lm3<-lm(fudata$logFUCL ~ fudata$DOSE1 + fudata$BSA1)
summary(red_lm3)
```



#### Forward model building is opposite; adds one at a time
#### Stepwise procedure does both at same time (software does this)

### How to detect/remove collinear predictors
#### find condition number (K)
```{r}
attach(fudata)
mm<-model.matrix(~+AGE+BSA+DOSE+SEX+MTX)
kappa(t(mm) %*% mm)

```

#### K is high, suggesting collinearity
### When scale predictors
```{r}
mm1<-model.matrix(~ + fudata$AGE1 + fudata$BSA1 + fudata$DOSE1 + fudata$SEX + fudata$MTX)
kappa(t(mm1) %*% mm1)

```

#### K dropped by 300,000-fold---
### #### Final model and scaled covariates--
```{r}
mm2<-model.matrix(~ + fudata$BSA1 + fudata$DOSE1)
kappa(t(mm2) %*% mm2)

```

### K dropped even further, now value suggests no collinearity--

#### Check Regression Diagnostics, assess normality of error assumptions--
```{r}
par(mfrow=c(2,2))
qqnorm(red_lm3$res,ylab="Raw Residuals",cex=1.5,cex.axis=1.5,cex.lab=1.5,pch=16)
qqline(red_lm3$res,lwd=2.0)
qqnorm(rstudent(red_lm3),ylab=" Studentized residuals",pch=16,cex=1.5,cex.axis=1.5,cex.lab=1.5)
abline(0,1,lwd=2.0)

```

### Draw histogram of residuals, should be around zero
```{r}
hist(red_lm3$res,10,cex=1.5,cex.axis=1.5,cex.lab=1.5,xlab="Raw Residuals",main="")

```

### Boxplot of residuals, should be around zero-

```{r}
boxplot(red_lm3$res,cex=1.5,cex.axis=1.5,cex.lab=1.5,pch=16,ylab="Raw Residuals")

```

#### Check constant variance assumption---
### Plot residual on y-axis, predicted log(CL) on x-axis, residuals should be zero--
```{r}
par(mfrow=c(1,2))
plot(fudata$logFUCL, red_lm3$res, cex=1.5,cex.axis=1.5,cex.lab=1.5,xlab="log(5FU CL)",
     ylab="Residuals")
abline(0,0,lwd=2.0)

```


#### Calculate leverage (influence in x-direction) with "hat function"--
```{r}
x<-model.matrix(red_lm3)
lev<-hat(x)

```

### plot leverage--
```{r}
par(mfrow=c(1,1))
plot(lev,ylab="Leverage",cex=2.0,cex.lab=1.5,cex.axis = 1.5,pch = 16)
abline(0.23,0, lwd=1.5)
names(lev)<-ID
lev[lev>0.23]

```

### Rule of thumb, if lev>2*p/n, then influential---

### Calculate standardized residuals (influence in y-directon)
### if residuals > +/- 2, then they're outliers
```{r}
par(mfrow=c(1,2))
plot(red_lm3$res,ylab="Residuals",cex=1.5,cex.lab=1.5,cex.axis = 1.5,pch= 16)

```

#### extract standardized outliers--
```{r}
plot(rstandard(red_lm3),ylab="StandardizedResiduals",cex=1.5,cex.lab=1.5,
     cex.axis = 1.5,pch = 16,ylim=c(-3,3))
abline(h=-2)
abline(h=2)

```

### Extract studentized outliers--
```{r}
plot(rstudent(red_lm3),ylab="Studentized Residuals",cex=1.5,cex.lab=1.5,cex.axis = 1.5,pch = 16,ylim=c(-3,3))
abline(h=-2)
abline(h=2)

```

#### check measures of fits
```{r}
influence.measures(red_lm3)
#### those rows (subjects) with astericks are outliers
```




#### Interpretation of Categorical Covariates

### bivariate analysis
### dose/response with dDBP
### add 40 random 0 or 1 (males=0, females=1)

```{r}
Sex<-rbinom(40,1,0.50)
data$SEX<-Sex
head(data)
Sex1<-factor(data$SEX)
attach(data)
drlm<-lm(resp ~ dose + Sex1)
summary(drlm)

plot(dose,resp,cex = 1.5,cex.axis=1.5,cex.lab=1.5, ylab=
       "dDBP(mmHg)",xlab="Dose(mg)", pch = 16)
abline(1.6093, -0.9855, col="blue", lwd=2.5)
(1.6093)+(-2.6376)
abline(-1.0283, -0.9855, col="red", lwd=2.5)
legend("topright", lwd=c(1,1),col=c("blue", "red"), 
  legend = c("Males", "Females"))

```

