---
title: "BinomialData"
output: html_notebook
---
### Activate required libraries
```{r}
list.of.packages <- c("epitools", "ggplot2")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
lapply(list.of.packages, require, character.only = TRUE)
# this should return 2 TRUE statements (one for each package)
# you can also lapply library instead of require
```

##### read in adverse event (nausea) dataset
```{r}
aedata<-read.csv("Nausea.csv", header=TRUE)
head(aedata)
summary(aedata)
```


### Bar graphs
```{r}

plot1<-ggplot(data=aedata, aes(x=factor(Nausea), fill=TRT)) +
  geom_bar(stat="count") +
  facet_grid(.~ TRT) +
  xlab("Occurrence of Nausea") +
  ylab("Count")
plot1

```

#### Plot attributes - axis labels, axis titles, background, legend options
```{r}
plot1+ theme(panel.background = element_rect(fill='white', colour='red'))+ 
  theme(axis.title.y = element_text(colour = 'black', size = 20))+ 
  theme(axis.text.y = element_text(size = 18))+ 
  theme(axis.title.x = element_text(colour = 'black', size = 20))+ 
  theme(axis.text.x = element_text(size = 18))+ 
  theme(plot.title = element_text(lineheight=.8, face="bold"))+ 
  theme(legend.position="top")+ 
  theme(legend.title=element_blank())+ 
  theme(legend.text = element_text(colour="blue", size = 18, face = "bold"))

```

### Cross-Tabulation (2x2) table for Nausea by Treatment Option
```{r}
table(Nausea=aedata$Nausea,Treatment=aedata$TRT)

```

### Cross-Tabulation (2x2) table for Nausea by Gender
```{r}
table(Nausea=aedata$Nausea, Gender=aedata$Gender, Treatment=aedata$TRT)

```


### Chi-squared test - Nausea by TRT
```{r}
ae.test<-chisq.test(aedata$Nausea, aedata$TRT)

```

### Display test results
```{r}
ae.test

```

### Display observed/expected results
```{r}
ae.test$observed
ae.test$expected

```

### Display chi squared test p-value
```{r}
ae.test$p.value

```

### Determine chi squared critical value for df=1
```{r}
qchisq(0.99,1)

```

#### any chi squared test statistic greater than 3.84 (with df=1) will reject Null
```{r}
qnorm(0.025)
qnorm(0.975)

```


### Second method (besides chi sqd) to determine differences in variables
### Confidence intervals for the difference in the two proportions (for two scenarios)
```{r}
prop.test(c(34,7), c(100,50))
```

### 95% CI does NOT contain zero, hence reject Ho (Null hypoth)

### Subset Treatment by Drug
```{r}
aedata_drug<-aedata[aedata$TRT=="Drug",]

```

### Create 2x2 contingency table by nausea and gender
### for 100 sub who got drug
```{r}
table(aedata_drug$Nausea, aedata_drug$Gender)
```

## determine relative risk (RR) of sample (SRR) as surrogate for population RR
## find asymptotic 95% CIs for SRR
```{r}

gentable<-table(aedata_drug$Nausea, aedata_drug$Gender)
col.total_F<-gentable[1,1]+gentable[2,1]
col.total_M<-gentable[1,2]+gentable[2,2]
p_female<-gentable[2,1]/col.total_F
p_male<-gentable[2,2]/col.total_M

```

## Calculate SRR
```{r}
RR<-p_female/p_male
RR

```

#### C.I for log(RR)
```{r}
SE_logRR<-sqrt((1-p_female)/(p_female*col.total_F)+
                 (1-p_male)/(p_male*col.total_M))
logRR<-log(p_female/p_male)
ci.lb<-logRR-1.96*SE_logRR
ci.ub<-logRR+1.96*SE_logRR

```

#### Exponentiate to get CI on normal scale
```{r}
ci_RR<-c(RR,exp(ci.lb),exp(ci.ub))
ci_RR

```

### Odds ratio (OR) is another way to determine associations b/w 2 nominal variables
```{r}
OR = (p_female/(1-p_female))/(p_male/(1-p_male))
OR
log_OR = log(OR)
log_OR

```

###### C.I for OR 
```{r}
SE_logOR<-sqrt((1/gentable[1,1])+(1/gentable[1,2])+(1/gentable[2,1]) + 
                 (1/gentable[2,2])) 
ci.lb<-log_OR - 1.96 * SE_logOR
ci.ub<-log_OR + 1.96 * SE_logOR

```

#### Exponentiate to get CI on normal scale
```{r}
ci_OR<-c(OR,exp(ci.lb),exp(ci.ub))
ci_OR

```

### Fisher's Exact Test (for small sample sizes) ###
```{r}
fisher.test(aedata_drug$Nausea, aedata_drug$Gender, alternative = "two.sided",
            conf.int = TRUE, conf.level = 0.95,
            simulate.p.value = TRUE, B = 100)

```

#### Different method for RR and OR
##### Look at risk ratio documentation
```{r}
?riskratio

```

### risk ratio template
```{r}
riskratio.wald(aedata_drug$Nausea, aedata_drug$Gender,
               conf.level = 0.95,
               rev = c("both"),
               correction = TRUE,
               verbose = FALSE)

```

##### Look at odds ratio documentation
```{r}
?oddsratio

```
#### Calculate odds ratio
```{r}
oddsratio.fisher(aedata_drug$Nausea, aedata_drug$Gender,
                 conf.level = 0.95,
                 rev = c("both"),
                 correction = FALSE,
                 verbose = FALSE)


```

