---
title: "Longitudinal"
output: html_notebook
---

## Read in and explore data
```{r}
resp<-read.csv("resp.csv", sep=",")
head(resp)

```
## Load required packages
```{r}
list.of.packages <- c("reshape2", "gee","geepack","lme4")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
lapply(list.of.packages, require, character.only = TRUE)
# this should return 4 TRUE statements (one for each package)
# you can also lapply library instead of require
```


###### Cross tabulations
```{r}
 ###just look at binary resp (nstat) by month
ftable(addmargins(xtabs(~nstat+month,data=resp)))
 ### now add treatment (drug or placebo) into the mix
ftable((xtabs(~treatment+nstat+month,data=resp)))
 ### add baseline resp as covariate
ftable((xtabs(~baseline+nstat+month,data=resp)))

```

### To identify potential predictor variables
###### Rearrange data for stacked bar chart

```{r}
count1<-ftable(xtabs(~treatment+nstat+month,data=resp))
```

### Melt fnc converts data from a wide format to a 
```{r}
mydf = melt(count1)
names(mydf) = c("TRT","Outcome","Visit","Count")

```

### Plot it
###### Stacked bar chart
```{r}
plot1<-ggplot(mydf, aes(x=Visit, y=Count, fill=Outcome )) +
  geom_bar(stat="identity")+
  facet_grid(.~TRT)
plot1+ theme(panel.background = element_rect(fill='white', colour='red'))+
  theme(axis.title.y = element_text(colour = 'black', size = 20))+
  theme(axis.text.y = element_text(size = 18))+
  theme(axis.title.x = element_text(colour = 'black', size = 20))+
  theme(axis.text.x = element_text(size = 18))+
  theme(plot.title = element_text(lineheight=.8, face="bold"))+
  theme(legend.position="top")+
  theme(legend.title=element_blank())+
  theme(legend.text = element_text(colour="blue", size = 18, face = "bold"))+
  theme(strip.text.x = element_text(size = 12, face = "bold",colour = "black"))

```

###### Fit a marginal model

### Base 
```{r}
resp_gee0<-gee(nstat ~ 1, data = resp, family = "binomial", id = subject,
               corstr = "independence", scale.fix = TRUE, scale.value = 1)
summary(resp_gee0)

```

### Full
```{r}
resp_gee1<-gee(nstat ~ centre + treatment + sex + baseline + age, 
      data = resp, family = "binomial", id = subject,
      corstr = "independence", scale.fix = TRUE, scale.value = 1)
summary(resp_gee1)
round(summary(resp_gee1)$coefficients,4)
round(summary(resp_gee1)$working.correlation,3)

```

### Independenet matrix not correct, try exchangeable
###### Exchangeable correlation
```{r}
resp_gee2<-gee(nstat ~ centre + treatment + sex + baseline + age, 
      data = resp, family = "binomial", id = subject, 
      corstr = "exchangeable", scale.fix = TRUE, scale.value = 1)
summary(resp_gee2)
round(summary(resp_gee2)$coefficients,4)
round(summary(resp_gee2)$working.correlation,3)

```

###### Hypothesis testing
```{r}
teststat.gee2<-round(summary(resp_gee2)$coefficients,4)
```

#### Calculate p-value
```{r}
pval<-round(pnorm(abs(teststat.gee2[,5]),lower.tail=FALSE)*2,4)
```

## Final table
```{r}
cbind(teststat.gee2,pval)

```

### Remove gender, age from model b/c not significant
### Final model
```{r}
resp_gee3<-gee(nstat ~ treatment + baseline, 
      data = resp, family = "binomial", id = subject, 
      corstr = "exchangeable", scale.fix = TRUE, scale.value = 1)
summary(resp_gee3)
round(summary(resp_gee3)$coefficients,4)
round(summary(resp_gee3)$working.correlation,3)

```

###### Hypothesis testing
```{r}
teststat.gee3<-round(summary(resp_gee3)$coefficients,4)
```

#### Calculate p-value
```{r}
pval3<-round(pnorm(abs(teststat.gee3[,5]),lower.tail=FALSE)*2,4)
```

## Final table
```{r}
cbind(teststat.gee3,pval3)
```

#### Robust SE very close to naive SE, thus correl matrix strucutre is adequate

### interpretation of population averages through odds ratios
###### Oddsratio and confidence interval

```{r}
se<-summary(resp_gee3)$coefficients["treatmenttreatment","Robust S.E."]
ci<-exp(coef(resp_gee3)["treatmenttreatment"] + c(-1,1) * se * qnorm(0.975))
cbind(OR=exp(coef(resp_gee3)["treatmenttreatment"]),ci)

```

##### odds ratio and CI for baseline as predictor---
###### Oddsratio and confidence interval---------------

```{r}
se1<-summary(resp_gee3)$coefficients["baselinepoor","Robust S.E."]
ci1<-exp(coef(resp_gee3)["baselinepoor"] + c(-1,1) * se1 * qnorm(0.975))
cbind(OR=exp(coef(resp_gee3)["baselinepoor"]),ci1)

```



#### Generalized linear mixed effects model (GLMM)---
```{r}
resp.glmm0<-glmer(nstat ~ 1 + (1|subject),
                data = resp, family = binomial, REML=F )
summary(resp.glmm0)

resp.glmm<-glmer(nstat ~ centre + treatment + baseline + sex + age +
      (1|subject), data = resp, family = binomial, REML=F )
summary(resp.glmm)

```

### removed age, gender from model to get final model--
```{r}
resp.glmm2<-glmer(nstat ~ centre + treatment + baseline+(1|subject),
                  data = resp, family = binomial, REML=F )
summary(resp.glmm2)

```

### removed age, gender, center from model to get final model--
```{r}
resp.glmm1<-glmer(nstat ~ treatment + baseline+ (1|subject),
    data = resp, family = binomial, REML=F )
summary(resp.glmm1)

```

### output full fixed effect data--
```{r}
exp(fixef(resp.glmm1))

```

#####Compare gee, glmm and glm fits------
```{r}
out.compare<-round(cbind(exp(coef(resp_gee3)), exp(fixef(resp.glmm2))),3)

colnames(out.compare)<-c('GEE','GLMM')
View(out.compare)

```

