---
title: "PM section - Pain Relief"
output: html_notebook
---
### Read in and explore data
```{r}
pain<-read.csv("Painrelief.csv", sep=",")
summary(pain)

```


### Load required packages
```{r}
list.of.packages <- c("PKNCA", 
                      "knitr",
                      "reshape2", 
                      "gee",
                      "geepack", 
                      "nlme", 
                      "lme4",
                      "lmtest",
                      "ggplot2",
                      "plotrix")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
lapply(list.of.packages, require, character.only = TRUE)
# this should return 10 TRUE statements (one for each package)
# you can also lapply library instead of require
```


##### Exploratory plots

##### Conc vs time curve, all together
```{r}
plot1<-ggplot(data = pain, aes(x = TIME, y = CONC, group=DOSE))+geom_point()+
  xlab("Time (hr)")+
  ylab("Concentration (ng/mL)")
plot1+theme(panel.background = element_rect(fill='grey', colour='red'))+
  theme(axis.title.y = element_text(colour = 'black', size = 20))+
  theme(axis.text.y = element_text(size = 18))+
  theme(axis.title.x = element_text(colour = 'black', size = 20))+
  theme(axis.text.x = element_text(size = 18))+
  theme(plot.title = element_text(lineheight=.8, face="bold"))+
  theme(legend.position="top")+
  theme(legend.title=element_blank())+
  theme(legend.text = element_text(colour="blue", size = 18, face = "bold"))

```


#### group data by DOSE and plot
```{r}
pain.2<-groupedData(CONC~TIME|DOSE, data=pain, 
     labels = list(x="Time(hr)", y="Conc(ng/mL)"))
plot2<-plot(pain.2, aspect=1/1, ylab="Concentration (ng/mL)", 
      xlab="Time (hr)")
plot2
plot22<-ggplot(pain, aes(displ, hwy)) + 
  geom_point() + 
  geom_line(data = grid, colour = "blue", size = 1.5) + 
  geom_text(data = pain, aes(label = model))

plot23<-ggplot(data=pain, aes(x=TIME, y=CONC))+
                 geom_point()

```


#### log-transform the drug concentration
```{r}
pain.2$logConc<-log(pain.2$CONC)
pain.2log<-groupedData(logConc~TIME|DOSE, data=pain.2, 
        labels = list(x="Time(hr)", y="Log Conc(ng/mL)"))
plot2.log<-plot(pain.2log, aspect=1/1, 
            ylab="Log Concentration (ng/mL)", 
            xlab="Time (hr)")

```

###### Plot Dose versus pain relief
```{r}
plot3<-ggplot(pain, aes(x = DOSE, y = PAINRELIEF)) +
  stat_sum(aes(size = ..n.., group = 1)) +
  scale_size_area(max_size=10)
plot3

```

######  Exploratory Response Plots
```{r}
## Create 2x2 contingency table by pain relief and Dose for all 160 sub-
table(pain$PAINRELIEF, pain$DOSE)

```

###### cross tabulations
```{r}
###just look at binary resp (nstat) by month--
ftable(addmargins(xtabs(~PAINRELIEF+TIME,data=pain)))

```

```{r}
### now add treatment (drug or placebo) into the mix---
ftable((xtabs(~DOSE+PAINRELIEF+TIME,data=pain)))

```

### To identify potential predictor variables
```{r}
###### Re arrange data for stacked bar chart

count1<-ftable(xtabs(~PAINRELIEF+DOSE,data=pain))
count1
```


```{r}
### melt fnc converts data from a wide format to a 
mydf = melt(count1)
names(mydf) = c("PainRelief", "Dose", "Count")
head(mydf)

```

###plot it
```{r}
###### Stacked bar chart
plot4<-ggplot(mydf, aes(x= Dose, y= Count, fill=PainRelief)) +
  geom_bar(stat="identity")
plot41<-ggplot(mydf, aes(x= Dose, y= Count, fill=PainRelief)) +
  geom_bar(stat="identity")
plot4+ theme(panel.background = element_rect(fill='gray', colour='red'))+
  theme(axis.title.y = element_text(colour = 'black', size = 20))+
  theme(axis.text.y = element_text(size = 18))+
  theme(axis.title.x = element_text(colour = 'black', size = 20))+
  theme(axis.text.x = element_text(size = 18))+
  theme(plot.title = element_text(lineheight=.8, face="bold"))+
  theme(legend.position="right")+
  theme(legend.title=element_blank())+
  theme(legend.text = element_text(colour="blue", size = 18, face = "bold"))+
  theme(strip.text.x = element_text(size = 12, face = "bold",colour = "black"))

```

##### count of pain relief by time


```{r}
count2<-ftable(xtabs(~PAINRELIEF+TIME,data=pain))
count2
mydf2 = melt(count2)
names(mydf2) = c("PainRelief", "Time", "Count")
mydf2

plot5<-ggplot(mydf2, aes(x= Time, y= Count, fill=PainRelief)) +
  geom_bar(stat="identity")

```

#### plot attributes - axis labels, axis titles, background, #####legend options
```{r}
plot5+ theme(panel.background = element_rect(fill='gray', colour='red'))+
  theme(axis.title.y = element_text(colour = 'black', size = 20))+
  theme(axis.text.y = element_text(size = 18))+
  theme(axis.title.x = element_text(colour = 'black', size = 20))+
  theme(axis.text.x = element_text(size = 18))+
  theme(plot.title = element_text(lineheight=.8, face="bold"))+
  theme(legend.position="right")+
  theme(legend.title=element_blank())+
  theme(legend.text = element_text(colour="blue", size = 18, face = "bold"))+
  theme(strip.text.x = element_text(size = 12, face = "bold",colour = "black"))

```

###### Using only successfully responses
### only PainRelief=1 responses
```{r}

nopain<-pain[pain$PAINRELIEF==1,]
count3<-ftable(xtabs(~PAINRELIEF+DOSE,data=nopain))

```

### melt fnc converts data from a wide to a narrow format  
```{r}

mydf3 = melt(count3)
names(mydf3) = c("PainRelief", "Dose", "Count")
head(mydf3)

```

###### Stacked bar chart
```{r}

ggplot(mydf3, aes(x= Dose, y= Count, fill=PainRelief)) +
  geom_bar(stat="identity")

```


##### count of pain relief by time
```{r}
count4<-ftable(xtabs(~PAINRELIEF+TIME,data=nopain))
mydf4 = melt(count4)
names(mydf4) = c("PainRelief", "Time", "Count")
ggplot(mydf4, aes(x= Time, y= Count, fill=PainRelief)) +
  geom_bar(stat="identity")
write.csv(mydf4, file="mydf4.csv")

```


#### Overlaid plots
```{r}
ggplot() + 
  geom_bar(data=mydf4, aes(x = Time, weight = Count))

ggplot() + 
  geom_bar(data=pain, aes(x = TIME, weight = PAINRELIEF)) +
  geom_point(data=pain, aes(x = TIME, y = DOSE, color = as.factor(PAINRELIEF))) +
  scale_y_continuous(sec.axis = sec_axis(~.*1, name = "Dose"))

ggplot() + 
  geom_bar(data=pain, aes(x = TIME, weight = PAINRELIEF)) +
  geom_point(data=pain, aes(x = TIME, y = CONC*10, color = as.factor(PAINRELIEF))) +
  scale_y_continuous(sec.axis = sec_axis(~./10, name = "Concentration"))

ggplot() + 
  geom_bar(data=pain, aes(x = TIME, weight = PAINRELIEF)) +
  geom_point(data=pain, aes(x = TIME, y = CONC*12,color = factor(PAINRELIEF), 
        shape = factor(DOSE))) +
  scale_y_continuous(sec.axis = sec_axis(~./12, name = "Concentration"))


plot45<-ggplot(mydf4, aes(x= Time, y= Count)) +
  geom_bar(stat="identity") + 
  geom_point(data=pain, aes(x=TIME, y=CONC), color='green')
plot45

```

## add extra space to right margin of plot within frame
```{r}
par(mar=c(5, 4, 4, 6) + 0.1)
```

### mar - A numeric vector of length 4, which sets the margin sizes in the 
##### following order: bottom, left, top, and right.
##### The default is c(5.1, 4.1, 4.1, 2.1)
## Plot first set of data and draw its axis
```{r}
plot(pain$TIME, pain$CONC, pch=16, axes=FALSE, ylim=c(0,10), xlab="", ylab="", 
     type="p",col="black")
axis(2, ylim=c(0,1),col="black",las=1)  ## las=1 makes horizontal labels
mtext("Drug Concentration (ng/mL)",side=2,line=2.5)
box()



## Allow a second plot on the same graph

par(new=TRUE)



## Plot the second plot and put axis scale on right

plot(mydf4$Time, mydf4$Count, pch=15,  xlab="", ylab="", ylim=c(0,120), 
     axes=FALSE, type="l", col="red")
## a little farther out (line=4) to make room for labels
mtext("Count of Pain Relief",side=4,col="red",line=4) 
axis(4, ylim=c(0,110), col="red",col.axis="red",las=1)

## Draw the time axis
axis(1,xlim=c(0,8))
axis(1,pretty(range(pain$TIME),1))
mtext("Time (Hours)",side=1,col="black",line=2.5)  

## Add Legend
legend("topleft",legend=c("Drug Conc","Pain Relief Count"),
       text.col=c("black","red"),pch=c(16,15),col=c("black","red"))

example(twoord.plot)
par(mar = c(5, 4, 4, 4) + 0.3)  # Leave space for z axis
plot(pain$TIME, pain$CONC) # first plot
par(new = TRUE)
plot(mydf4$Time, mydf4$PainRelief, type = "l", axes = FALSE, bty = "n",
     ylim=c(0,120), xlab = "", ylab = "")
axis(side=4)
mtext("z", side=4, line=3)

```


###### NCA analysis on pain data to get PK values ###

### Perform NCA


## By default it is groupedData; convert it to a data frame for use--
### maps conc to any variable or grouping factor ##
```{r}
my.conc<-PKNCAconc(as.data.frame(pain), CONC~TIME|ID)
```


## Dosing data needs to only have one row per dose, so subset for that first
```{r}
d.dose<-unique(pain[pain$TIME == 0.0,
                       c("ARM","ID","TIME","CONC", "PAINRELIEF","DOSE")])

knitr::kable(d.dose,
             caption="Dosing data extracted from data set")

```

### By dose
```{r}
my.dose <- PKNCAdose(d.dose, DOSE~TIME|ID)
my.dose
```

#### Combine dose and conc data
```{r}
my.data.automatic <- PKNCAdata(my.conc, my.dose)
summary(my.data.automatic)
knitr::kable(PKNCA.options("single.dose.aucs"))
knitr::kable(my.data.automatic$intervals)

```

### Compute parameters
```{r}
my.results.automatic <- pk.nca(my.data.automatic)
summary(my.results.automatic)

```

### Specify start/stop times
```{r}
my.intervals <- data.frame(start=0.0,
                           end=Inf,
                           cmax=TRUE,
                           tmax=TRUE,
                           aucinf=TRUE,
                           auclast=TRUE,
                           cl=TRUE,
                           vz=TRUE,
                           half.life=TRUE)

my.data.manual <- PKNCAdata(my.conc, my.dose,intervals=my.intervals)
knitr::kable(my.data.manual$interval)
summary(my.data.manual)

my.results.manual <- pk.nca(my.data.manual)
summary(my.results.manual)
my.results.manual$result

write.csv(my.results.manual$result, file="PKNCAdata.csv")
knitr::kable(summary(my.results.manual))

```

### Make a data.frame for PK manual results
```{r}
PKresults<-my.results.manual$result
```

#### Drop first two columns in PK results
```{r}
PKresults1<-PKresults[c(-1, -2)]
```

#### Subset out only t=0hr rows in main data.frame
```{r}
pain.0<-pain[pain$TIME==0,]

PKresults2<-t(PKresults1)

```

### Melt PK data
```{r}
mPKresults<-melt(PKresults1, id=c("ID", "PPTESTCD"))
```

### Cast the melted data
### Cast(data, formula, function)
```{r}
PKresults3<-dcast(mPKresults, ID~variable+PPTESTCD)
write.csv(PKresults3, file="PKNCAresults.csv")
```

#### Cbind pk results to subsetted data.frame
```{r}
pain.nca<-cbind2(PKresults3, pain.0)
head(pain.nca)
write.csv(pain.nca, file="FinalPKNCA.csv")
```

### Drop columns except cmax, auclast, HL
```{r}
pain.nca1<-pain.nca[c(-2, -4,-7,-8,-9,-10,-11,-12,-13,-14,-15,-16,-17,-18,
                      -19,-20,-21,-22,-23,-24,-25,-27)]
write.csv(pain.nca1, file="FinalPKNCA1.csv")
head(pain.nca1)
class(pain.nca1$PPORRES_auclast)<- "numeric"
round(pain.nca1$PPORRES_auclast,3)

nca<-read.csv("FinalPKNCA1.csv", header=TRUE)

```

### Plot overlay of dose vs AUC, with dose vs PainRelief to show that increasing exposure does NOT increase response
```{r}
boxplot(PPORRES_auclast ~ DOSE, data = nca, main="AUClast by Dose", xlab="Dose (mg)",
        ylab="AUClast (hr*ng/mL)")
boxplot(PPORRES_cmax ~ DOSE, data = nca, main="Cmax by Dose", xlab="Dose (mg)",
        ylab="Cmax (ng/mL)")

```



#### Generalized linear mixed models for longitudinal binary responses

###### Fit a marginal model-----------------------
```{r}
pain_gee1 <- gee(PAINRELIEF ~ as.factor(DOSE)+CONC+TIME, 
    data = pain, family = "binomial", id = ID, 
    corstr = "independence", scale.fix = TRUE, scale.value = 1)
summary(pain_gee1)

```

###### Hypothesis testing
```{r}
teststat.gee1<-round(summary(pain_gee1)$coefficients,4)
```

#### Calculate p-value
```{r}
pval1<-round(pnorm(abs(teststat.gee1[,5]),lower.tail=FALSE)*2,4)
```

## Final table
```{r}

cbind(teststat.gee1,pval1)
```


### Independenet matrix not correct, try exchangeable

###### Exchangeable correlation
```{r}
pain_gee2<- gee(PAINRELIEF ~ as.factor(DOSE)+CONC+TIME, data = pain, 
    family = "binomial", id = ID, corstr = "exchangeable", 
    scale.fix = TRUE, scale.value = 1)
summary(pain_gee2)
round(summary(pain_gee2)$coefficients,4)
round(summary(pain_gee2)$working.correlation,3)

```

#### Auto-regressive correl matrix
```{r}
pain_gee3<-gee(PAINRELIEF~ as.factor(DOSE)+CONC+TIME, data = pain, 
    family="binomial",  id=ID, corstr="AR-M", 
    scale.fix=TRUE, scale.value=1)
summary(pain_gee3)
round(summary(pain_gee3)$coefficients,4)
round(summary(pain_gee3)$working.correlation,3)

```

#### Unstructured matrix
```{r}
pain_gee4<-gee(PAINRELIEF~as.factor(DOSE)+CONC+TIME, data=pain, 
    family="binomial", id=ID, corstr="unstructured", 
    scale.fix=TRUE, scale.value=1)
summary(pain_gee4)   
round(summary(pain_gee4)$coefficients,4)
round(summary(pain_gee4)$working.correlation,3)

```

#### Looks the best

###### Full Model Hypothesis Testing
```{r}
teststat.gee4<-round(summary(pain_gee4)$coefficients,4)
```

#### Calculate p-value
```{r}

pval4<-round(pnorm(abs(teststat.gee4[,5]),lower.tail=FALSE)*2,4)

```

## Final table
```{r}
cbind(teststat.gee4,pval4)

```


#### Baseline model; intercept-only model
```{r}
pain_geeBL<-gee(PAINRELIEF~1, data=pain, family="binomial", 
    id=ID, corstr="unstructured", scale.fix=TRUE, scale.value=1)
summary(pain_geeBL)
round(summary(pain_geeBL)$coefficients,4)
round(summary(pain_geeBL)$working.correlation,3)

```

##### Odds ratio and CI for baseline as predictor
```{r}
se.BL <- summary(pain_geeBL)$coefficients["(Intercept)","Robust S.E."]
ci.BL<-exp(coef(pain_geeBL)["(Intercept)"] + c(-1,1) * se.BL * qnorm(0.975))
cbind(OR=exp(coef(pain_geeBL)["(Intercept)"]),ci.BL)

```

###### Base Model Hypothesis Testing
```{r}
teststat.geeBL<-round(summary(pain_geeBL)$coefficients,5)
```

#### Calculate p-value
```{r}
pval.BL<-pnorm(abs(teststat.geeBL[,5]),lower.tail=FALSE)*2
```

## Final table
```{r}
cbind(teststat.geeBL,pval.BL)

baselineprob=exp(-0.5402513)/(1+exp(-0.5402513))
baselineprob

```


### Interpretation of population averages through odds ratios
###### Oddsratio and confidence inetrval 
```{r}
se.dose5 <- summary(pain_gee4)$coefficients["as.factor(DOSE)5","Robust S.E."]
ci.dose5<-exp(coef(pain_gee4)["as.factor(DOSE)5"] + c(-1,1) * se.dose5 * qnorm(0.975))
cbind(OR=exp(coef(pain_gee4)["as.factor(DOSE)5"]),ci.dose5)

se.dose20<- summary(pain_gee1)$coefficients["as.factor(DOSE)20","Robust S.E."]
ci.dose20<-exp(coef(pain_gee1)["as.factor(DOSE)20"] + c(-1,1) * se.dose20 * qnorm(0.975))
cbind(OR=exp(coef(pain_gee1)["as.factor(DOSE)20"]),ci.dose20)

se.dose80<- summary(pain_gee1)$coefficients["as.factor(DOSE)80","Robust S.E."]
ci.dose80<-exp(coef(pain_gee1)["as.factor(DOSE)80"] + c(-1,1) * se.dose80 * qnorm(0.975))
cbind(OR=exp(coef(pain_gee1)["as.factor(DOSE)80"]),ci.dose80)

```

#### Generalized linear mixed effects model (GLMM)

```{r}
DOSE0520<-pain$DOSE[pain$DOSE<80]
DOSE0520

```

#### Baseline model (intercept-only)
```{r}
pain.glmmBL<-glmer(PAINRELIEF ~ 1+(1|ID),data = pain, family = binomial, REML=F)
pain.glmmBL
summary(pain.glmmBL)

```

#### FULL model
```{r}
pain.glmm<-glmer(PAINRELIEF ~ as.factor(DOSE)+TIME+CONC+(1|ID),data = pain, 
        family = binomial, REML=F)
summary(pain.glmm)
```

###### Test the significance of adding ALL covariates = Full model vs base


## loglikelihood ratio test (2*(LLfull-LLbase/red))
```{r}
lrt.BLFull = 2*((-1020.7) - (-1089.131))
lrt.BLFull
```

### Determine chi squared critical value for df1
```{r}
qchisq(0.95,5)
lrtest(pain.glmm, pain.glmmBL)

```

### Added random effect to CONC and ID
```{r}
pain.glmm1<-glmer(PAINRELIEF ~ as.factor(DOSE)+TIME+CONC+(1|ID)+(1|CONC),
        data = pain, family = binomial, REML=TRUE)
summary(pain.glmm1)
lrtest(pain.glmm1, pain.glmm)

```

### output full fixed effect data
```{r}
exp(fixef(pain.glmm1))
```


### Added random effect to CONC and ID, REMOVED 'TIME'
```{r}
pain.glmm2<-glmer(PAINRELIEF ~ as.factor(DOSE)+CONC+(1|ID)+(1|CONC),
          data = pain, family = binomial, REML=F)
summary(pain.glmm2)
exp(fixef(pain.glmm2))
lrtest(pain.glmm1, pain.glmm2)

```


### Removed CONC and RE on CONC
```{r}
pain.glmm3<-glmer(PAINRELIEF ~ as.factor(DOSE)+(1|ID),data = pain, family = binomial, REML=F)
summary(pain.glmm3)

```

##### Compare gee, glmm and glm fits
```{r}
out.compare <- round(cbind(exp(coef(pain_gee4)), exp(fixef(pain.glmm1))),3)
rownames(out.compare) <- names(coef(pain_gee4))
colnames(out.compare) <- c('GEE','GLMM')
out.compare

```

