---
title: "R Notebook"
output: html_notebook
---

### Read in and view summary of data
```{r}
pain<-read.csv("Painrelief.csv", sep=",")
painHL<-read.csv("Painrelief_modHL.csv", sep=",")
remed<-read.csv("TimetoRemed.csv", sep=",")
summary(pain)
```


### Load required packages
```{r}
list.of.packages <- c("reshape2", 
                      "gee",
                      "geepack", 
                      "lme4",
                      "lmtest",
                      "ggplot2",
                      "survival")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
lapply(list.of.packages, require, character.only = TRUE)
# this should return 7 TRUE statements (one for each package)
# you can also lapply library instead of require
```



##### Exploratory plots-
##### Conc vs time curve, all together
```{r}
library(ggplot2)
plot <- ggplot(data = pain, aes(x = TIME, y = CONC, group=ARM))+geom_point()+
  xlab("Time (hr)")+
  ylab("Concentration (ng/mL)")
plot
```

##### Pain Relief by Dose
```{r}

plot1<-ggplot(data=pain, aes(x=PAINRELIEF, fill=DOSE)) + 
  geom_bar(stat="bin")+
  facet_grid(. ~ DOSE)+
  xlab("Pain Relief")+
  ylab("Count")
plot1+theme(panel.background = element_rect(fill='grey', colour='red'))+
  theme(axis.title.y = element_text(colour = 'black', size = 20))+
  theme(axis.text.y = element_text(size = 18))+
  theme(axis.title.x = element_text(colour = 'black', size = 20))+
  theme(axis.text.x = element_text(size = 18))+
  theme(plot.title = element_text(lineheight=.8, face="bold"))+
  theme(legend.position="top")+
  theme(legend.title=element_blank())+
  theme(legend.text = element_text(colour="blue", size = 18, face = "bold"))

plot2<-ggplot(data=pain, aes(x=PAINRELIEF,fill=TIME)) + 
  geom_bar(stat="bin")+
  facet_grid(. ~ TIME)+
  xlab("Pain Relief")+
  ylab("Count")
plot2+theme(panel.background = element_rect(fill='grey', colour='red'))+
  theme(axis.title.y = element_text(colour = 'black', size = 20))+
  theme(axis.text.y = element_text(size = 18))+
  theme(axis.title.x = element_text(colour = 'black', size = 20))+
  theme(axis.text.x = element_text(size = 18))+
  theme(plot.title = element_text(lineheight=.8, face="bold"))+
  theme(legend.position="top")+
  theme(legend.title=element_blank())+
  theme(legend.text = element_text(colour="blue", size = 18, face = "bold"))
par(mfrow=c(1,2))
plot1
plot2

```

```{r}
plot3<-ggplot(pain, aes(x = DOSE, y = PAINRELIEF)) +
  stat_sum(aes(size = ..n.., group = 1)) +
  scale_size_area(max_size=10)
plot3

```


## Create 2x2 contingency table by pain relief and Dose for all 160 subjects

```{r}
table(pain$PAINRELIEF, pain$DOSE)
table(painHL$PAINRELIEF, painHL$CONC)

```


#### Generalized linear mixed models for longitudinal binary responses

###### Cross tabulations
### Just look at binary resp (nstat) by month
```{r}
ftable(addmargins(xtabs(~PAINRELIEF+TIME,data=pain)))
```

### now add treatment (drug or placebo) into the mix
```{r}
ftable((xtabs(~DOSE+PAINRELIEF+TIME,data=pain)))
```


### To identify potential predictor variables
###### Re arrange data for stacked bar chart
```{r}
count1<-ftable(xtabs(~PAINRELIEF+TIME,data=pain))
count1
count2<-ftable(xtabs(~PAINRELIEF+DOSE,data=pain))
count2

```

### melt fnc converts data from a wide format to a 

```{r}
mydf = melt(count1)
names(mydf) = c("PainRelief", "Time", "Count")
head(mydf)
mydf2 = melt(count2)
names(mydf2) = c("PainRelief", "Dose", "Count")
mydf2

```


###plot it
###### Stacked bar chart
```{r}
plot4<-ggplot(mydf, aes(x= Time, y= Count, fill=PainRelief)) +
  geom_bar(stat="identity")
plot4

```

#### plot attributes - axis labels, axis titles, background, #####legend options

```{r}
plot4+ theme(panel.background = element_rect(fill='gray', colour='red'))+
  theme(axis.title.y = element_text(colour = 'black', size = 20))+
  theme(axis.text.y = element_text(size = 18))+
  theme(axis.title.x = element_text(colour = 'black', size = 20))+
  theme(axis.text.x = element_text(size = 18))+
  theme(plot.title = element_text(lineheight=.8, face="bold"))+
  theme(legend.position="right")+
  theme(legend.title=element_blank())+
  theme(legend.text = element_text(colour="blue", size = 18, face = "bold"))+
  theme(strip.text.x = element_text(size = 12, face = "bold",colour = "black"))

plot5<-ggplot(mydf2, aes(x= Dose, y= Count, fill=PainRelief)) +
  geom_bar(stat="identity")

```

#### Plot attributes - axis labels, axis titles, background, legend options
```{r}
plot5+ theme(panel.background = element_rect(fill='gray', colour='red'))+
  theme(axis.title.y = element_text(colour = 'black', size = 20))+
  theme(axis.text.y = element_text(size = 18))+
  theme(axis.title.x = element_text(colour = 'black', size = 20))+
  theme(axis.text.x = element_text(size = 18))+
  theme(plot.title = element_text(lineheight=.8, face="bold"))+
  theme(legend.position="right")+
  theme(legend.title=element_blank())+
  theme(legend.text = element_text(colour="blue", size = 18, face = "bold"))+
  theme(strip.text.x = element_text(size = 12, face = "bold",colour = "black"))

```

###### Fit a marginal model

```{r}
pain_gee1 <- gee(PAINRELIEF ~ as.factor(DOSE)+CONC+TIME, data = pain, family = "binomial", id = ID, corstr = "independence", scale.fix = TRUE, scale.value = 1)
summary(pain_gee1)
```

###### Hypothesis testing
```{r}
teststat.gee1<-round(summary(pain_gee1)$coefficients,4)
```

#### calculate p-value 
```{r}
pval1<-round(pnorm(abs(teststat.gee1[,5]),lower.tail=FALSE)*2,4)
```

## final table-
```{r}
cbind(teststat.gee1,pval1)
```

### independenet matrix not correct, try exchangeable
###### Exchangeable correlation
```{r}
pain_gee2<- gee(PAINRELIEF ~ as.factor(DOSE)+CONC+TIME, data = pain, family = "binomial", id = ID, corstr = "exchangeable", scale.fix = TRUE, scale.value = 1)
summary(pain_gee2)
round(summary(pain_gee2)$coefficients,4)
round(summary(pain_gee2)$working.correlation,3)

```

#### Auto-regressive correl matix
```{r}
pain_gee3<-gee(PAINRELIEF~DOSE+CONC+TIME,data=pain, family="binomial", id=ID, corstr="AR-M", scale.fix=TRUE, scale.value=1)
summary(pain_gee3)

```

#### Unstructured matrix
```{r}
pain_gee4<-gee(PAINRELIEF~as.factor(DOSE)+CONC+TIME, data=pain, family="binomial", id=ID, corstr="unstructured", scale.fix=TRUE, scale.value=1)
summary(pain_gee4)   ####Looks the best-

```

#### Baseline model; intercept-only model
```{r}
pain_geeBL<-gee(PAINRELIEF~1, data=pain, family="binomial", id=ID, corstr="unstructured", scale.fix=TRUE, scale.value=1)
summary(pain_geeBL)
baselineprob=exp(-0.5402513)/(1+exp(-0.5402513))
baselineprob
```

###### Hypothesis testing
```{r}
teststat.gee4<-round(summary(pain_gee4)$coefficients,4)
```

#### calculate p-value 
```{r}
pval4<-round(pnorm(abs(teststat.gee4[,5]),lower.tail=FALSE)*2,4)
```

## final table-
```{r}
cbind(teststat.gee4,pval4)
```

  
### interpretation of population averages through odds ratios
###### Oddsratio and confidence interval 
```{r}
se.dose5 <- summary(pain_gee1)$coefficients["as.factor(DOSE)5","Robust S.E."]
ci.dose5<-exp(coef(pain_gee1)["as.factor(DOSE)5"] + c(-1,1) * se * qnorm(0.975))
cbind(OR=exp(coef(pain_gee1)["as.factor(DOSE)5"]),ci.dose5)

se.dose20<- summary(pain_gee1)$coefficients["as.factor(DOSE)20","Robust S.E."]
ci.dose20<-exp(coef(pain_gee1)["as.factor(DOSE)20"] + c(-1,1) * se * qnorm(0.975))
cbind(OR=exp(coef(pain_gee1)["as.factor(DOSE)20"]),ci.dose20)

se.dose80<- summary(pain_gee1)$coefficients["as.factor(DOSE)80","Robust S.E."]
ci.dose80<-exp(coef(pain_gee1)["as.factor(DOSE)80"] + c(-1,1) * se * qnorm(0.975))
cbind(OR=exp(coef(pain_gee1)["as.factor(DOSE)80"]),ci.dose80)

```

##### odds ratio and CI for baseline as predictor
```{r}
se.BL <- summary(pain_geeBL)$coefficients["(Intercept)","Robust S.E."]
ci.BL<-exp(coef(pain_geeBL)["(Intercept)"] + c(-1,1) * se * qnorm(0.975))
cbind(OR=exp(coef(pain_geeBL)["(Intercept)"]),ci.BL)
```



#### Generalized linear mixed effects model (GLMM)

```{r}
DOSE0520<-pain$DOSE[pain$DOSE<80]
DOSE0520

```

#### Baseline model (intercept-only)
```{r}
pain.glmmBL<-glmer(PAINRELIEF ~ 1+(1|ID),data = pain, family = binomial, REML=F)
pain.glmmBL
summary(pain.glmmBL)
```

#### FULL model
```{r}
pain.glmm<-glmer(PAINRELIEF ~ as.factor(DOSE)+TIME+CONC+(1|ID),data = pain, family = binomial, REML=F)
summary(pain.glmm)
```

###### Test the significance of adding ALL covariates = Full model vs base 


## Log likelihood ratio test (2*(LLfull-LLbase/red)) 
```{r}
lrt.BLFull = 2*((-1020.7) - (-1089.131))
lrt.BLFull
```

### Determine chi squared critical value for df1
```{r}
qchisq(0.95,5)
```


### Added random effect to CONC and ID
```{r}
pain.glmm1<-glmer(PAINRELIEF ~ as.factor(DOSE)+TIME+CONC+(1|ID)+(1|CONC),data = pain, family = binomial, REML=F)
summary(pain.glmm1)
lrt1=2*((-972.4) - (-1020.7))
lrt1
```

### output full fixed effect data
```{r}
exp(fixef(pain.glmm1))
```


### Added random effect to CONC and ID, REMOVED 'TIME'
```{r}
pain.glmm2<-glmer(PAINRELIEF ~ as.factor(DOSE)+CONC+(1|ID)+(1|CONC),data = pain, family = binomial, REML=F)
summary(pain.glmm2)
exp(fixef(pain.glmm2))
lrt2=2*((-972.4) - (-973.2))
lrt2
```

### Removed CONC and RE on CONC
```{r}
pain.glmm3<-glmer(PAINRELIEF ~ as.factor(DOSE)+(1|ID),data = pain, family = binomial, REML=F)
summary(pain.glmm3)

```

#####Compare gee, glmm and glm fits
```{r}
## Need to have resp_gee3 from Day3_longitudinal
out.compare <- round(cbind(exp(coef(resp_gee3)), exp(fixef(resp.glmm1))),3)
rownames(out.compare) <- names(coef(resp_gee3)) 
colnames(out.compare) <- c('GEE','GLMM')
out.compare

```


#####  Continuation of Day 3's PM case study ######

```{r}
remed<-read.csv("TimetoRemed.csv", sep=",")
head(remed)
```

##### K-M curve (Probability of remedication by Dose)

```{r}
km.fit1<-survfit( Surv(REMEDTime, REMEDStatus)~ARM, data=remed) 
summary(km.fit1)
plot(km.fit1,lty=c(1,3),cex=1.8,cex.lab=1.5,cex.axis=1.5, 
     main="Probability of Remedication by Dose", xlab="Time (hr)",
     ylab="Probability of Remedication", lwd=3.0, 
     col=c("red","blue", "green", "black"))
legend("bottomleft",c("80mg","20 mg", "5mg", "Placebo"),lty=c(1,3),
       col=c("green","red", "blue", "black"), 
       cex=1.0,lwd=1.0, bty="n")
abline(h=0.5)
```



#### Cumulative Hazard fnc; remed by dose
```{r}
plot(km.fit1,fun="cumhaz",lty=c(1,3),cex=1.8,cex.lab=1.5,
     cex.axis=1.5, xlab="Time (hr)",ylab="Cumulative Hazard", 
     lwd=3.0, col=c("red","blue", "green", "black"))
legend("topleft",c("80mg","20 mg", "5mg", "Placebo"),lty=c(1,3),
       col=c("green","red", "blue", "black"), 
       cex=1.0,lwd=2.0, bty="n")
```



### Bind AUC exposure data to remed data ##
```{r}
## Use the FinalPKNCA1.csv created by Day3_pmsection.R or Day3_PMpain.Rmd
nca<-read.csv("FinalPKNCA1.csv", header=TRUE)

remed2<-cbind(remed,nca$PPORRES_auclast)
remed3<-cbind(remed2, nca$DOSE)

remed4<-within(remed3,{DOSE<-factor(ARM,
    labels=c(0,5,20,80),levels=c("Placebo", "A5_0_at2h", 
      "A20_0_at2h", "A80_0_at2h"))})
```

### COX MODEL - intercept only
```{r}
cph.int<- coxph(Surv(REMEDTime, REMEDStatus) ~ 1,
                data = remed3,ties="breslow") 
summary(cph.int)
```


#### FULL Model ###
```{r}
cph.full<- coxph(Surv(REMEDTime, REMEDStatus) ~ DOSE + 
    nca$PPORRES_auclast, data = remed4, ties="breslow") 
summary(cph.full)
```


#### Reduced Model w/o AUC
```{r}
cph.red<- coxph(Surv(REMEDTime, REMEDStatus) ~ DOSE, 
      data = remed4, ties="breslow") 
print(cph.red)
summary(cph.red)
class(cph.red)

```



###### Test the significance of placebo 
```{r}
names(remed4)
cph4.noplaceb <- coxph(Surv(REMEDTime, REMEDStatus) ~ nca$PPORRES_auclast, data = remed4, ties="breslow") 
summary(cph4.noplaceb)
cph4.noplaceb$loglik

```

## Partial loglikelihood ratio test 
```{r}
lrtest(cph.full, cph.red)
```

