---
title: "R Notebook"
output: html_notebook
---

### Read in data

```{r}
data(ovarian)
head(ovarian)
```


### Load required packages
```{r}
list.of.packages <- c("survminer",
                      "survival")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
lapply(list.of.packages, require, character.only = TRUE)
# this should return 7 TRUE statements (one for each package)
# you can also lapply library instead of require
```


#### Summarize data
```{r}
ovarian<-within(ovarian,{
  rx<-factor(rx)
  ecog.ps<-factor(ecog.ps)
  resid.ds<-factor(resid.ds)
  fustat<-factor(fustat)})
summary(ovarian)
```


###### Fit Kaplan-Meier  
 
```{r}
km.fit<-survfit( Surv(futime, fustat)~1, data=ovarian) 
km.fit
plot(km.fit, cex=1.8, cex.lab=1.5, cex.axis=1.5, xlab="Time, days", 
     ylab="Probability of survival", lwd=2.0, col="red")

```

### Summarize to show data points that are plotted on K-M plot
```{r}
summary(km.fit)
```


### Compare survival between treatments
```{r}
km.fit1<-survfit(Surv(futime, fustat)~rx, data=ovarian)
km.fit1 
summary(km.fit1)
```

### Plot the survival curves from each treatment
```{r}
plot(km.fit1, lty=c(1,3),cex=1.8,cex.lab=1.5,cex.axis=1.5, 
     xlab="Time, days",ylab="Probability of survival", lwd=3.0, 
     col=c("red","blue")) 
legend("bottomleft",c("Rx=1","Rx=2"),lty=c(1,3),col=c("red","blue"), 
       cex=1.5,lwd=2.0, bty="n") 
```


#### Cumulative hazard function  
```{r}
plot(km.fit1,fun="cumhaz",lty=c(1,3),cex=1.8,cex.lab=1.5,cex.axis=1.5, 
     xlab="Time, days",ylab="Probability of survival", lwd=3.0, col=c("red","blue"))
```


### Are survival distribs different between treatment groups?
### Use hypothesis testing using log rank test
```{r}
class(ovarian$rx)<-"numeric"
class(ovarian$fustat)<-"numeric"
test<-survdiff(Surv(futime, fustat)~rx,data = ovarian) 
test 
```

### p-value not sig, thus no diff, Ho not rejected

### Test whether age groups affect survival
### Mathangi  needs to provide code she used to make age groups
```{r}
test2<-survdiff(Surv(futime, fustat)~agecut,data = ovarian) 
test2
```
#### Cox models

```{r}
cph.fit <- coxph(Surv(futime, fustat) ~ 
                   rx+resid.ds+ecog.ps+age,data = ovarian,ties="breslow") 
summary(cph.fit) 
```

###### Test the significance of age 
```{r}
cph.fit.noage <- coxph(Surv(futime, fustat) ~ rx+resid.ds+ecog.ps,data = ovarian, ties="breslow") 
cph.fit.noage$loglik
cph.fit$loglik
```

## Partial log likelihood ratio test 
```{r}
lrt.age = 2*(cph.fit$loglik[2] - cph.fit.noage$loglik[2]) 
lrt.age
```


###### Final model  
```{r}
cph.fit.age<-coxph(Surv(futime, fustat) ~ age, data = ovarian,method="breslow")
summary(cph.fit.age)
```


##### Estimated Survival function
```{r}
newdf=data.frame(age=c(55,60,70))
plot(survfit(cph.fit.age,newdf),cex=1.8,cex.lab=1.5, cex.axis=1.5, xlab="Time,days",lty=c(1,2,3), ylab="Estimated Probability of survival", lwd=3.0)
legend("bottomleft", legend=c("Age=55yrs","Age=60yrs", "Age=70yrs"),lty=c(1,2,3)) 

```

##### Check proportional odds assumption for age as predictor 
```{r}
agecut<-cut(ovarian$age,c(quantile(ovarian$age,0:2/2))) 
kmfit.age<-survfit(Surv(futime, fustat)~agecut, data=ovarian) 
plot(kmfit.age, lty=c(1,2),cex=1.8,cex.lab=1.5,cex.axis=1.5, xlab="Time, days",ylab="Probability of survival", lwd=3.0, col=c("red","blue"))
legend("bottomleft",c("Age<=57","Age>57" ),lty=c(1,2,3),col=c("red","blue"), cex=1.5, lwd=2.0, bty="n")
```


#### Types of Residuals 
```{r}
mtg.1 = resid(cph.fit.age,type="martingale") 
coxsn.1 = ovarian$fustat - mtg.1 
dev.1 = resid(cph.fit.age,type="deviance") 
score.1 = resid(cph.fit.age,type="score") 
schoen.1 = resid(cph.fit.age,type="schoenfeld") 
dfb.1 = resid(cph.fit.age,type="dfbeta") 
dfbs.1 = resid(cph.fit.age,type="dfbetas") 
```


##### Cox proportional hazards assumption 
```{r}
temp <- cox.zph(cph.fit.age) 
temp 
plot(temp) 
```


#### Overall fit - Cox Snell residual 
```{r}
cumhaz_coxsnell<-survfit(Surv(coxsn.1,ovarian$fustat)~1,type="fleming-harrington") 
plot(cumhaz_coxsnell$time,-log(cumhaz_coxsnell$surv),xlim=c(0,3), ylim=c(0,3),xlab="Cox-snell residual", ylab="Cumulative hazard of CS-residuals", main="Overall fit of the model",cex=1.5,cex.axis=1.5,cex.lab=1.5)
lines(c(0,3),c(0,3),lwd=2.0) 
```


#### Deviance residual-Outlier detection 
```{r}
plot(dev.1,ylab="Deviance residual",
     type="b",cex=1.5,cex.axis=1.5,cex.lab=1.5, 
     xlab="Subject number", ylim=c(-3,3)) 
abline(h=0, lty=2, lwd = 2.0, col="blue") 
plot(ovarian$age,dev.1,ylab="",cex=1.5,cex.axis=1.5,cex.lab=1.5, xlab="Age, years", ylim=c(-3,3)) 
lines(lowess(ovarian$age,dev.1)) 
abline(h=0, lty=2, lwd = 2.0, col="blue")

```

### Influential points 
```{r}
par(mfrow=c(1,1))
fb.1 = resid(cph.fit.age,type="dfbeta") 
plot(dfb.1,cex.lab=1.4,cex.axis=1.5, cex=2.0,type="b") 
abline(h=0, lty=2, lwd = 2.0, col="blue") 
```