---
title: "R Notebook"
output: html_notebook
---

#Goal
Plot data from two different dataframes onto same plot, each with separate y-axes, but same x-axis time scale (0-8hrs).

#Data
Two dataframes
- The first is x vs y, time vs drug concentration, from 5mg, 20mg or 80mg doses (painrelief.csv).
- The second is count data of pain relief (yes/no) at each time point, subsetted out only for successes (=1) (mydf4.csv).  
#Plot 
Count data as a barplot along same time scale (x-axis), overlaid with the data points of drug conc.

#Setup
Create painPlot folder in working directory and put painrelief.csv and mydf4.csv inside.

## Read in the data
```{r}
path<- paste(getwd(), "/painPlot", sep = "")
setwd(path)
painrelief <- read.csv("painrelief.csv")
mydf4 <- read.csv("mydf4.csv")
```

## Install and load ggplot (if not installed)

```{r}
#require returns (invisibly) a logical indicating whether the required package is available.
if(!require("ggplot2")){ 
    install.packages("ggplot2")
    library(ggplot2)
}

```
## mydf4 barplot

```{r}
ggplot() + 
  geom_bar(data=mydf4, aes(x = Time, weight = Count))
 
```
## Plot painrelief barplot as above and Dose vs. Time
```{r}
ggplot() + 
  geom_bar(data=painrelief, aes(x = TIME, weight = PAINRELIEF)) +
  geom_point(data=painrelief, aes(x = TIME, y = DOSE, color = as.factor(PAINRELIEF))) +
  scale_y_continuous(sec.axis = sec_axis(~.*1, name = "Dose"))

```
## Plot painrelief barplot as above and Concentration vs. Time
```{r}
ggplot() + 
  geom_bar(data=painrelief, aes(x = TIME, weight = PAINRELIEF)) +
  geom_point(data=painrelief, aes(x = TIME, y = CONC*10, color = as.factor(PAINRELIEF))) +
  scale_y_continuous(sec.axis = sec_axis(~./10, name = "Concentration"))

```
## Add ARM as shape and adjust 2nd y axis
```{r}
ggplot() + 
  geom_bar(data=painrelief, aes(x = TIME, weight = PAINRELIEF)) +
  geom_point(data=painrelief, aes(x = TIME, y = CONC*12, color = factor(PAINRELIEF), shape = factor(ARM))) +
  scale_y_continuous(sec.axis = sec_axis(~./12, name = "Concentration"))

```