---
title: "R Notebook"
output: html_notebook
---

#Goal
Plot data from two different dataframes onto same plot, each with separate y-axes, but same x-axis time scale (0-8hrs).

#Data
Two dataframes
- The first is x vs y, time vs drug concentration, from 5mg, 20mg or 80mg doses (painrelief.csv).
- The second is count data of pain relief (yes/no) at each time point, subsetted out only for successes (=1) (mydf4.csv).  
#Plot 
Count data as a barplot along same time scale (x-axis), overlaid with the data points of drug conc.

#Setup
Create painPlot folder in working directory and put painrelief.csv and mydf4.csv inside.

## Read in the data
```{r}
path<- paste(getwd(), "/painPlot", sep = "")
setwd(path)
painrelief <- read.csv("painrelief.csv")
mydf4 <- read.csv("mydf4.csv")
```

## Install and load ggplot (if not installed)
```{r}
list.of.packages <- c("ggplot2", "Rcpp")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
```

```{r}
#require returns (invisibly) a logical indicating whether the required package is available.
if(!require("ggplot2")){ 
    install.packages("ggplot2")
    library(ggplot2)
}

```


Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).
